The Role of Ice Compositions for Snowlines and the C/N/O Ratios in Active Disks

### 2016-07-21 tex/draft2/ms_revised.tex
upd

Both in Solar system comets and in protoplanetary disks, volatile carbon and oxygen are primarily contained in H$_2$O, CO$_2$ and CO (e.g., \citealt{lodders03}, \citealt{mumma11}, \citealt{oberg11}, \citealt{boogert15}). However, some fraction of carbon may also be carried by CH$_4$ (e.g., \citealt{oberg08}), which may change the C/O ratio in gas and in dust at some disk locations. In the case of nitrogen, chemical models of the protostellar nebula (e.g., \citealt{owen01}) and of protoplanetary disks (e.g., \citealt{rodgers02}) suggest that N$_2$ was the dominant form of nitrogen, and that giant planets have accreted their nitrogen content primarily as N$_2$ \citep{mousis14}. Observations of Solar system bodies such as Titan and Pluto show that N$_2$ is prevalent in their atmospheres (\citealt{cruikshank93}, \citealt{owen93}), \textbf{although Titan's N$_2$ dominated atmospheric composition is not primordial but rather originates from primordial NH$_3$ accreted by the satellite (\citealt{atreya78}, \citealt{mandt14}).}  %Moreover, the Rosetta spacecraft has recently made the first direct measurement of the N$_2$ abundance in comet 67P/Churyumov-Gerasimenko \citep{rubin15}. 
%\textbf{While CO and N$_2$ have similar condensation temperatures (e.g., \citealt{fayolle16}), their snowline locations may }  
Due to the high volatility of N$_2$, the gas phase nitrogen-to-oxygen (N/O) ratio in the outer disk may be even more enhanced than the C/O ratio compared to its average value in the disk. \textbf{The larger N/O ratio enhancement is due to the fact that nitrogen is primarily carried by N$_2$, while carbon has other carriers besides CO. The N/O enhancement is most pronounced between the CO and N$_2$ snowlines, where the gas phase N/O ratio exceeds the stellar value by many orders of magnitude due to the depletion of oxygen gas in this region (see Section \ref{sec:results}).} Giant planets that form at wide separations should thus have an excess of nitrogen in their atmospheres, which could be used to trace their formation origin. In addition to N$_2$, a fraction of the nitrogen abundance may also be carried by less volatile species such as NH$_3$ (\citealt{bottinelli10}, \citealt{mumma11}).

The snowline locations of the main carbon, oxygen and nitrogen carriers strongly depend on the ice grain \textbf{composition}. Very volatile species, such as CO and N$_2$, present binding energies, and therefore snowline locations, that are sensitive to the details of the \textbf{composition} of the icy grain mantles. 
\textbf{Spectroscopic observations suggest that up to 90\% of the CO ice is frozen in a layer that is thick enough and separated from the H$_2$O ice layer underneath so that it can be considered pure ice (e.g., \citealt{pontoppidan03}). However, if the CO (or N$_2$) ice layer is thin enough ($\sim$monolayer coverage), then it will interact with the H$_2$O ice substrate (e.g., \citealt{collings03}). The ice binding energy is significantly larger in this water dominated environment than in the pure ice case, as shown by laboratory experiments (\citealt{collings03}, \citealt{oberg05}, \citealt{bisschop06}, \citealt{fayolle16}).} %have shown that CO and N$_2$ have significantly different binding energies depending on whether they are pure or water dominated ices. 
This implies that ices in different environments will sublimate at different radii, which will substantially change the disk regions where these volatiles are present in gaseous or solid form (see Section \ref{sec:static}).  \textbf{In protostellar cores, H$_2$O ice is primarily amorphous (e.g., \citealt{williams02}, \citealt{vandishoeck14}), and can either be compact or porous.  When temperatures exceed $\sim$80-90 K, H$_2$O ice acquires a crystalline structure \citep{schegerer10}. The CO binding energy is larger in an amorphous porous H$_2$O ice environment than in the amorphous compact or crystalline cases (\citealt{noble12}, \citealt{fayolle16}). The N$_2$ binding energy is also larger for a porous versus compact H$_2$O ice substrate. No equivalent studies have yet been performed for the deposition of N$_2$ on crystalline H$_2$O ice, but we expect the N$_2$ binding energy in this environment to follow the same trend as the CO binding energy, as CO and N$_2$ display a similar desorption behavior. To explore the range of distances at which CO and N$_2$ in different environments desorb, we consider the limiting cases: pure ices (lowest binding energy) and ices residing on an amorphous porous H$_2$O ice substrate (highest binding energy). We refer to the latter simply as water dominated ices unless noted otherwise.}

>>>>

Both in Solar system comets and in protoplanetary disks, volatile carbon and oxygen are primarily contained in H$_2$O, CO$_2$ and CO (e.g., \citealt{lodders03}, \citealt{mumma11}, \citealt{oberg11}, \citealt{boogert15}). However, some fraction of carbon may also be carried by CH$_4$ (e.g., \citealt{oberg08}), which may change the C/O ratio in gas and in dust at some disk locations. In the case of nitrogen, chemical models of the protostellar nebula (e.g., \citealt{owen01}) and of protoplanetary disks (e.g., \citealt{rodgers02}) suggest that N$_2$ was the dominant form of nitrogen, and that giant planets have accreted their nitrogen content primarily as N$_2$ \citep{mousis14}. %Observations of Solar system bodies such as Titan and Pluto show that N$_2$ is prevalent in their atmospheres (\citealt{cruikshank93}, \citealt{owen93}), \textbf{although Titan's N$_2$ dominated atmospheric composition is not primordial but rather originates from primordial NH$_3$ accreted by the satellite (\citealt{atreya78}, \citealt{mandt14}).}  %Moreover, the Rosetta spacecraft has recently made the first direct measurement of the N$_2$ abundance in comet 67P/Churyumov-Gerasimenko \citep{rubin15}. 
%\textbf{While CO and N$_2$ have similar condensation temperatures (e.g., \citealt{fayolle16}), their snowline locations may }  
Due to the high volatility of N$_2$, the gas phase nitrogen-to-oxygen (N/O) ratio in the outer disk \textbf{is expected to be high, perhaps} more enhanced than the C/O ratio compared to the Solar value. %\textbf{The larger N/O ratio enhancement is due to the fact that nitrogen is primarily carried by N$_2$, while carbon has other carriers besides CO. The N/O enhancement is most pronounced between the CO and N$_2$ snowlines, where the gas phase N/O ratio exceeds the stellar value by many orders of magnitude due to the depletion of oxygen gas in this region (see Section \ref{sec:results}).} 
\textbf{The C/O ratio in gas cannot exceed unity (i.e., a factor of $\sim$2 enhancement compared to the Solar value) since the major volatile carbon carrier is CO. In contrast, the N/O ratio mainly depends on the relative depletion of N$_2$ and oxygen carriers, and it will increase as each of the oxygen carrier snowlines (H$_2$O, CO$_2$, CO) is crossed. Beyond the CO snowline, there is no strict upper limit to the N/O ratio. The spatial extent of this latter region depends on the relative bond strengths of CO and N$_2$ to ice, but may be quite large (see Section \ref{sec:results}).} Giant planets that form at wide separations should thus have an excess of nitrogen in their atmospheres, which could be used to trace their formation origin. In addition to N$_2$, a fraction of the nitrogen abundance may also be carried by less volatile species such as NH$_3$ (\citealt{bottinelli10}, \citealt{mumma11}). \textbf{The present day N$_2$ in Titan's atmosphere, for example, is thought to originate from accretion of primordial NH$_3$ (\citealt{atreya78}, \citealt{mandt14}).}

The snowline locations of the main carbon, oxygen and nitrogen carriers strongly depend on the ice grain \textbf{composition}. Very volatile species, such as CO and N$_2$, present binding energies, and therefore snowline locations, that are sensitive to the details of the \textbf{composition} of the icy grain mantles. 
\textbf{Spectroscopic observations suggest that up to 90\% of the CO ice is frozen in a layer that is thick enough and separated from the H$_2$O ice layer underneath so that it can be considered pure ice (e.g., \citealt{pontoppidan03}). However, if the CO (or N$_2$) ice layer is thin enough ($\sim$monolayer coverage), then it will interact with the H$_2$O ice substrate (e.g., \citealt{collings03}). The ice binding energy is significantly larger in this water dominated environment than in the pure ice case, as shown by laboratory experiments (\citealt{collings03}, \citealt{oberg05}, \citealt{bisschop06}, \citealt{fayolle16}).} %have shown that CO and N$_2$ have significantly different binding energies depending on whether they are pure or water dominated ices. 
This implies that ices in different environments will sublimate at different radii, which will substantially change the disk regions where these volatiles are present in gaseous or solid form (see Section \ref{sec:static}).  \textbf{In protostellar cores, H$_2$O ice is primarily amorphous (e.g., \citealt{williams02}, \citealt{vandishoeck14}).  When temperatures exceed $\sim$80-90 K, H$_2$O ice acquires a crystalline structure \citep{schegerer10}. The CO binding energy is larger in an amorphous porous H$_2$O ice environment than in the amorphous compact or crystalline cases (\citealt{noble12}, \citealt{fayolle16}). The N$_2$ binding energy is also larger for a porous versus compact H$_2$O ice substrate. No equivalent studies have yet been performed for the deposition of N$_2$ on crystalline H$_2$O ice, but we expect the N$_2$ binding energy in this environment to follow the same trend as the CO binding energy, as CO and N$_2$ display a similar desorption behavior. To explore the range of distances at which CO and N$_2$ in different environments desorb, we consider the limiting cases: pure ices (lowest binding energy) and ices residing on an amorphous porous H$_2$O ice substrate (highest binding energy). We refer to the latter simply as water dominated ices unless noted otherwise.}
